### 2016-01-26 inputs/discussion.tex
modifiche 26 mattina 10:36

\subsection{Role of crop species: rice vs. alfalfa}
\label{sec:role_of_species}
Imputation results were significantly different in rice and alfalfa: all imputation algorithms performed consistently better in rice data sets, where Beagle, RFI and KNNI achieved performances comparable to what is reported in literature for SNP-chip data (e.g.,~\cite{vanraden_genomic_2013} for bovine data;
%problems with this citation: url contains hashes
\cite{the_1000_genomes_project_consortium_integrated_2012}
for human data). On the other hand, imputation accuracy in alfalfa data sets was much lower. This can be ascribed to four main factors: 
i) the imputation problem was simpler in rice than in alfalfa, as the rice dataset comprised only pure lines, i.e. only two genotype classes instead of alfalfa's three; \\
ii) the genome structure was simpler in rice (diploid) than alfalfa (autotetraploid, rendered diploid ``in silico'' during SNP calling);\\
iii) alfalfa datasets did not contain population structure \cite{Annicchiarico2015}, while rice population was intrinsically stratified - being a collection of five subpopulation. Population structure is implicitly used by algorithms as Beagle and KNNI that create clusters of similar genotypes. Thus, rice dataset population structure facilitated the inputation process;\\
iv) rice data were analyzed by chromosome, allowing for higher average values of linkage disequilibrium (LD) among SNP markers, thus simplifying the imputation process.\\
Finally, the further difference in imputation accuracy between Alfalfa-Med and Alfalfa-PV suggests that other factors, directly connected to GBS protocols, may influence the general quality of the resulting data.

@@

\subsection{Problem size}
\label{sec:size_of_problem}
Problem size, besides affecting the accuracy achievable by some imputation methods, is mainly relevant with reference to the required computation resources. RFI was the most demanding algorithm, with computation times growing exponentially both with the total number of marker genotypes (m SNP $\times$ n samples) and with the number of missing genotypes to be imputed. When analyzing the complete rice dataset (all chromosomes together), RFI took a maximum of about 40 days to complete imputation, and became computationally intractable (on the available platform) for the largest missing rate scenarios. This was partially mitigated by parallelization (we used 10 CPUs in our experiments). All other imputation algorithms took much shorter times to complete imputation, even in the most challenging scenarios. Putting together imputation accuracy and computation time, we found that the best performing imputation algorithms were Beagle in rice and KNNI in alfalfa.\\
The analysis of the rice complete data set (437 samples $\times$ 166\,418 SNP markers) offered interesting insights on the applicability of imputation methods. While problems related to the computational requirements can usually be tackled by the constant improvement in bioinformatics facilities, current trends in genotyping suggest that imputation methods and computational strategies well suited for large data sets are to be preferred. The non-linearity of computational efficiency generated, for most examined imputation algorithms, a total computation time for the complete rice data set which was greater than the sum of the computation times required by the individual chromosomes. Again, RFI provided the most blatant illustration of this behavior, with computation times for the complete rice data set going up to 937 hours (about 40 days), while the sum of the computation times for all the single chromosomes in the same experimental conditions added to 88.3 hours.

>>>>

\subsection{Imputation efficiency: differences in rice vs. alfalfa}
\label{sec:role_of_species}
Imputation results were significantly different in rice and alfalfa: all imputation algorithms performed consistently better in rice data sets, where Beagle, RFI and KNNI achieved performances comparable to what is reported in literature for SNP-chip data (e.g.,~\cite{vanraden_genomic_2013} for bovine data;
%problems with this citation: url contains hashes
\cite{the_1000_genomes_project_consortium_integrated_2012}
for human data). On the other hand, imputation accuracy in alfalfa data sets was much lower. This can be ascribed to four main factors: 
i) the imputation problem was simpler in rice than in alfalfa, as the rice dataset comprised only pure lines, i.e. only two genotype classes instead of alfalfa's three; \\
ii) the genome structure was simpler in rice (diploid) than alfalfa (autotetraploid, rendered diploid ``in silico'' during SNP calling);\\
iii) alfalfa datasets did not contain population structure \cite{Annicchiarico2015}, while rice population was intrinsically stratified - being a collection of five subpopulation. Population structure is implicitly used by algorithms as Beagle and KNNI that create clusters of similar genotypes. Thus, rice dataset population structure facilitated the inputation process;\\
iv) rice data were analyzed by chromosome, allowing for higher average values of linkage disequilibrium (LD) among SNP markers, thus simplifying the imputation process.\\
Finally, the further difference in imputation accuracy between Alfalfa-Med and Alfalfa-PV suggests that other factors, directly connected to GBS protocols, may influence the general quality of the resulting data.

@@

\subsection{Size of the imputation problem}
\label{sec:size_of_problem}
Albeit this was not the main objective of the paper, and no formal effort to optimise the implementation of the various imputation methods was done, still the recorded computation times give us some interesting information.
The size of the problem, besides affecting the accuracy achievable by some imputation methods, is mainly relevant with reference to the required computation resources. RFI was the most demanding algorithm, with computation times growing exponentially both with the total number of marker genotypes (m SNP $\times$ n samples) and with the number of missing genotypes to be imputed. When analyzing the complete rice dataset (all chromosomes together), RFI took a maximum of about 40 days to complete imputation, and became computationally intractable (on the available platform) for the largest missing rate scenarios. This was partially mitigated by parallelization (we used 10 CPUs in our experiments). All other imputation algorithms took much shorter times to complete imputation, even in the most challenging scenarios. Putting together imputation accuracy and computation time, we found that the best performing imputation algorithms were Beagle in rice and KNNI in alfalfa.\\

Problem size, besides affecting the accuracy achievable by some imputation methods, is mainly relevant with reference to the required computation resources. RFI was the most demanding algorithm, with computation times growing exponentially both with the total number of marker genotypes (m SNP $\times$ n samples) and with the number of missing genotypes to be imputed. When analyzing the complete rice dataset (all chromosomes together), RFI took a maximum of about 40 days to complete imputation, and became computationally intractable (on the available platform) for the largest missing rate scenarios. This was partially mitigated by parallelization (we used 10 CPUs in our experiments). All other imputation algorithms took much shorter times to complete imputation, even in the most challenging scenarios. Putting together imputation accuracy and computation time, we found that the best performing imputation algorithms were Beagle in rice and KNNI in alfalfa.\\

The analysis of the rice complete data set (437 samples $\times$ 166\,418 SNP markers) offered interesting insights on the applicability of imputation methods. While problems related to the computational requirements can usually be tackled by the constant improvement in bioinformatics facilities, current trends in genotyping suggest that imputation methods and computational strategies well suited for large data sets are to be preferred. The non-linearity of computational efficiency generated, for most examined imputation algorithms, a total computation time for the complete rice data set which was greater than the sum of the computation times required by the individual chromosomes. Again, RFI provided the most blatant illustration of this behavior, with computation times for the complete rice data set going up to 937 hours (about 40 days), while the sum of the computation times for all the single chromosomes in the same experimental conditions added to 88.3 hours.
